Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-15 manuscript/manuscript.Rmd
Study 2 methods & results

---
title: "Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination"
shorttitle        : "Flattering Advice"
author: 
  - name          : "Amanda Chen"
    affiliation   : "1"
    corresponding : yes    
    email         : "zchengj@connect.ust.hk"
    address       : "Department of Management, The Hong Kong University of Science and Technology, Hong Kong"
  - name          : "David Hagmann"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "The Hong Kong University of Science and Technology"
note : |
  `r format(Sys.time(), "%e %B, %Y")`
authornote : |
  The authors are grateful to a lot of things.
abstract: |
  The purpose of advice is to help recipients make better decisions and solve their problems. However, in this paper, we propose that advisors may also take recipients' responses into account and attempt to avoid disappointing them, leading to what we call "flattering advice." In two pre-registered experiments involving real interactions between advisors and advisees (n = 2,700), we show that individuals consider others' expectations when giving advice, even when such expectations do not provide informative insights for the underlying decision-making problem. As a result, men receive more aspirational advice than women, given their higher confidence, when expectations (but not gender) are shown to advisors. However, incorporating expectations leads to worse outcomes (Study 1). Individuals inflate their advice due to interpersonal considerations: people who are concerned about advisees' likability inflate their advice to a great extent. Advisees reward such flattering advice, finding the advisers more likable and no less trustworthy than those who provide more accurate advice (Study 2). We discuss practical and theoretical contributions and future research directions.

keywords          : "Gender difference, Advice, Interpresonal relationship"
wordcount         : "4,362"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
header-includes:
   - \usepackage{setspace, appendix, placeins, booktabs, tikz, siunitx}
   - \usepackage{tabularx, epigraph, float, colortbl, tabu, pdflscape}
   - \usetikzlibrary{positioning}
   - \raggedbottom
   - \usepackage[maxfloats=256]{morefloats}
   - \maxdeadcycles=1000
bibliography      : bibliography.bib
output: 
  # papaja::apa6_docx
  papaja::apa6_pdf
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

@@

<!-- David: Just commenting on what I tried to do here... a bit of a transition -->
<!-- for why this study follows from the last. -->
So far, we have shown that displaying expectations can induce gender
differences in advice given. Specifically, advisers were more likely to
recommend competing against a group of High Performers to men who had higher
expectations about their performance. We now test directly whether this 
is driven by interpersonal concerns. Specifically, we examine whether 
more flattering advice is the result of a desire to be liked and whether
advisees indeed do like advisers more when they give flattering advice.

@@

We recruited 300 participants and invited them to upload photos of
themselves (selfies) to be rated by other participants on
attractiveness. We ended up with a sample of
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women who agreed to do so and uploaded usable pictures. We selected the
first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years; 50.0% Female).
We informed them that their selfies will be randomly grouped into nine
others from the same gender and ranked by people of the opposite gender
and asked them to predict their rank, with 1st rank represents the most
attractive and 10th rank represents the least attractive. We assigned
each participant to a group of 10 participants of their own gender.

Next, we recruited `r scales::comma(dim(study2_time2)[1])`new
participants and asked them to rank the attractiveness of one such group
of 10 participants of the opposite gender ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). After
they did so, we showed them the photo of the participant they ranked as
the 7th most attractive (that is, the 4th least attractive) and reminded
them that they had ranked this person at 7th attractive. We (truthfully)
informed them that this participant would have a chance to earn a bonus
if they accurately guessed their rank based on the aggregated ratings of
all participants who had ranked this group. Participants were then
randomly assigned to an "accuracy" or a "likeability" treatment. In the
accuracy treatment, they were told that they would receive a bonus (\$1)
identical to that of the advisee. While in the "likeability" treatment,
the amount their bonus depended on how the advisee rated them in terms
of likeability, after observing only the rank that they had advised the
advisee to guess. Advisees would rate to what extent they think the
adviser is likable on a 5-point scale, and each point they earn
increases their bonus by 20 cents.

Finally, we invited participants from Stage 1 and were ranked as the 7th
attractive by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we kept the
survey open for 7 days. In total,
`r scales::comma(dim(study2_time3)[1])` participants
(`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])`
men,
`r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])`
women) returned. They were reminded of the selfie they uploaded in Stage
1 and informed that a group of 10 selfies, including theirs, had been
rated by other participants from Prolific. They were also introduced to
the opportunity to earn a \$1 bonus if they guessed their rank
accurately. Participants were then shown the advice from a randomly
selected adviser and made their predictions. After deciding which rank
to bet on, advisees were reminded of the advice they received and asked
to rate the advice giver's likability, warmth, friendliness,
good-naturedness, trustworthiness, and sincerity on a 5-point scale. The
last 6 items were adopted from @Fiske2007.

@@

```{r study2advice, fig.cap="Advice for 7th most attractive participants"}
library(dplyr)
library(ggplot2)

@@

# Create a facet plot with condition as the facet variable and gender as the line color
ggplot(advice_avg, aes(x = Advice, y = percent, color = Condition)) +
  geom_line(linewidth = 0.75)+
  scale_color_manual(values = c("#FFA500", "#4169E1"),
                      labels = c("Accurate", "Likable")) +
  scale_y_continuous(name = "% of participants") +
  scale_x_continuous(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

Again, we found supportive evidence that males are more confident about
themselves, giving higher initial rank prediction (more attractive) than
female participants
(`r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)`
vs.
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`,
`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`).

Examining advice from Stage 2 participants, we found that participants
advised a lower rank (that is, more attractive) in the likeability
treatment than the accuracy treatment
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`
vs
`r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
As seen in Figure \@ref(fig:study2advice), notably, however, even in the
accuracy treatment, participants advised a lower rank than they
themselves had provided
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that advisers inferred that advising someone that they
were more attractive would make the adviser appear more likeable and
therefore provided advice that communicated a more favorable impression
of the participants' attractiveness.

@@

To test whether flattering advice brings more positive evaluation of
advisers, we create a single index for warmth and trustworthiness by
taking average of ratings on warmth, friendliness, good-naturedness,
trustworthiness for the former and ratings on trustworthiness, and
sincerity for the later. As seen in Figure \@ref(fig:study2rating), we
found that advisees who suggested that the advisee was more attractive
(lower rank) were indeed rated as more likeable and warm (b = -0.109, p
\< .05; b = -0.133, p \< .001; See Columns 1 and 2 of Table 2).
Interestingly, these benefits are not at the cost of being hypocritical;
advisers are also seen as more trustworthy, but the result is only
directional (Column 3 of Table 2). Notably, however, we were not powered
to do a comparison across the two experimental groups.

# Discussion

dvice substantially shapes people's career and personal outcomes.
Therefore, understanding advisers' motivations has far-reaching
implications. We propose that advisers' focus on advisees' expectations
and the desire to avoid disappointing them can induce gender differences
in the advice given and ultimately in the outcomes achieved. Gender
differences in career outcomes contribute significantly to the gender
wage gap, with women often holding less senior positions and
experiencing worse career outcomes, even with equal performance.
Therefore, identifying a factor such as expectations that contributes to
this gap has important implications for organizational practice.

@@

The findings of this study suggest that organizations can promote gender
equity by calibrating employees' expectations or shifting advisers'
attention away from them. While these measures may not entirely solve
the gender gap problem, they can make a meaningful difference when
unconscious bias interventions have failed to show any effect
[@Chang2019; @Paluck2009]. We believe our findings may be particularly
relevant to societies that value signaling in interpersonal
relationships, including Hong Kong and many Asian countries. Future work
could also examine whether people in different cultures place more
weight on expectations.

>>>>

---
title: "Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination"
shorttitle        : "Flattering Advice"
author: 
  - name          : "Amanda Chen"
    affiliation   : "1"
    corresponding : yes    
    email         : "zchengj@connect.ust.hk"
    address       : "Department of Management, The Hong Kong University of Science and Technology, Hong Kong"
  - name          : "David Hagmann"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "The Hong Kong University of Science and Technology"
note : |
  `r format(Sys.time(), "%e %B, %Y")`
abstract: |
  The purpose of advice is to help recipients make better decisions and solve their problems. However, in this paper, we propose that advisors may also take recipients' responses into account and attempt to avoid disappointing them, leading to what we call "flattering advice." In two pre-registered experiments involving real interactions between advisors and advisees (n = 2,700), we show that individuals consider others' expectations when giving advice, even when such expectations do not provide informative insights for the underlying decision-making problem. As a result, men receive more aspirational advice than women, given their higher confidence, when expectations (but not gender) are shown to advisors. However, incorporating expectations leads to worse outcomes (Study 1). Individuals inflate their advice due to interpersonal considerations: people who are concerned about advisees' likability inflate their advice to a great extent. Advisees reward such flattering advice, finding the advisers more likable and no less trustworthy than those who provide more accurate advice (Study 2). We discuss practical and theoretical contributions and future research directions.

keywords          : "Gender difference, Advice, Interpresonal relationship"
wordcount         : "4,362"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa7"
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
classoption       : "man"
header-includes:
   - \usepackage{setspace, appendix, placeins, booktabs, tikz, siunitx}
   - \usepackage{tabularx, epigraph, float, colortbl, tabu, pdflscape}
   - \usetikzlibrary{positioning}
   - \raggedbottom
   - \usepackage[maxfloats=256]{morefloats}
   - \maxdeadcycles=1000
bibliography      : bibliography.bib
output: 
  # papaja::apa6_docx
  papaja::apa6_pdf
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

@@

<!-- David: Just commenting on what I tried to do here... a bit of a transition -->
<!-- for why this study follows from the last. -->
So far, we have shown that displaying expectations can induce gender
differences in advice-giving. Specifically, advisers were more likely to
recommend competing against a group of High Performers to men who had higher
expectations about their performance than did women. We now test directly whether this 
is driven by interpersonal concerns. Specifically, we examine whether 
more flattering advice is the result of a desire to be liked and whether
advisees indeed do like advisers more when they give flattering advice.

@@

We recruited 300 participants from Prolific and invited them to upload photos of
themselves (selfies) to be rated by other participants on
attractiveness.
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women agreed to do so and uploaded pictures that adhered to our
instructions (e.g., did not include other people). We selected the
first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years).
Participants were informed that their selfies would be randomly grouped with
those of nine other participants of their gender and ranked in terms of attractiveness
by a group of new Prolific participants of the opposite gender.
They then guessed their rank (unincentivized) and the survey concluded
with demographic questions **David: or did we start with those?**

Next, we recruited a new, gender-balanced sample for the role
of advisers. `r scales::comma(dim(study2_time2)[1])`
participants from Prolific ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started by
providing demographic information, then were matched to a group of the opposite
gender. They then ranked participants from most to least attractive using a
drop-down menu next to each picture.
<!-- David: wasn't there something with participants selecting the same rank more -->
<!-- than once, and we dropped people who did that? Or were we able to solve this -->
<!-- and it happened only during a pilot? -->
After submitting their ratings, they saw the photo of the participant they
had ranked as the 7th most attractive (i.e., the 4th least attractive). They
were reminded of the rank they had given to that perso and informed
that this participant would be invited back and could earn a \$1 bonus if they
guessed their rank correctly. The rank was determined by the aggregate ratings
of all participants who had ranked this group. Their task was to give advice
to this participant about the rank they should bet on based on having observed
all ten selfies and their own assessment. We randomly assigned participants to
one of two incentivization schemes. In the "Accuracy" treatment, they received
a bonus identical to the advisee: \$1 if they guessed their rank correctly.
In the "Likeability" treatment, we informed them that they would be rated by
the advisee on a 5-point Likert scale for how likeable they thought they were.
Each point on the scale would translate to a bonus of 20 cents. They then
selected a rank that they would recommend the advisee to bet on.

Finally, we invited participants from Stage 1 and were ranked as the 7th
attractive by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we kept the
survey open for 7 days. In total,
`r scales::comma(dim(study2_time3)[1])` participants
(`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])`
men,
`r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])`
women) returned. They were reminded of the selfie they uploaded in Stage
1 and informed that a group of 10 selfies, including theirs, had been
rated by other participants from Prolific. They then saw the advice from a 
randomly selected adviser and made their estimate with a \$1 incentive for
guessing accurately. They then saw the rank they had been advised to bet on
one more time and were asked
to rate the adviser's likability, warmth, friendliness,
good-naturedness, trustworthiness, and sincerity on 5-point Likert scales
(adapted from @Fiske2007).

@@

<!-- David: The treatment differences here are hard to see. The most salient feature -->
<!-- is the spike at 7th (which suggests no flattering advice), and the spike at 5. -->
<!-- Can we try a violin plot for the advised rank?  And maybe we can add a -->
<!-- dotted horizontal line at 7. This is would be consistent with what they said -->
<!-- just on the previous screen. So a good reference line. -->

```{r study2advice, fig.cap="Advice given to the 7th most attractive participants"}
library(dplyr)
library(ggplot2)

@@

# Create a facet plot with condition as the facet variable and gender as the line color
ggplot(advice_avg, aes(x = Advice, y = percent, color = Condition)) +
  geom_line(linewidth = 0.75)+
  scale_color_manual(values = c("#FFA500", "#4169E1"),
                      labels = c("Accurate", "Likable")) +
  scale_y_continuous(name = "% of participants") +
  scale_x_continuous(name = "Advised Rank",
                     breaks = seq(1, 10)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

We begin by examining the prior beliefs of advisees who uploaded their selfies.
On average, men guessed they ranked `r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)` in their group of 10 and women guessed that they ranked
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`.
Participants overall underestimate their attractiveness relative to the benchmark
average of 5.5. Moreover, women do so more than men
`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`.
Notably, people's self-perceptions correlated strongly/weakly with the aggregate
ratings of the advisers (correlation coefficient & cor.test). However, there
was substantial heterogeneity in perceptions of attractiveness.
Of the 200 participants, XXX were ranked as 7th most attractive by at least one
adviser. On average,
men in this subset estimated they were ranked Xth and women estimated they were
ranked Xth (t-test).

<!-- David: Some stats to fill in above. I think this would be interesting to include -->
<!-- even though we don't make any predictions about this. -->

Next, we turn our attention to the advisers (Figure \@ref(fig:study2advice)). 
In the Accuracy condition, those
uploading selfies were advised to bet on rank `r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`.
Notably, this is significantly more attractive than the 7th rank those advisers
had themselves guessed just on the prior screen
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that even when incentivized for accuracy, participants offered
flattering advice.^[This shading could be due to concerns of avoiding disappointment. However, it could also be that advisers are uncertain about the rankings they had given and make a recommendation that combines their own belief with a uniform prior.]
Importantly, and as predicted, we find that advisers in the Likeable treatment
recommend betting on a lower rank, communicating that they think the participant
in the selfie is more attractive
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
This suggests that advisers inferred that advising someone that they
were more attractive would make the adviser appear more likeable and
therefore provided advice that communicated a more favorable impression
of the participants' attractiveness. The distribution shown in Figure \@ref(fig:study2advice)
shows that participants do not simply tell participants that they are the
most attractive person in the group. They may infer that flattering advise needs
to be somewhat realistic to be believable. We return to this in the general discussion.

<!-- David: Making a note here that we should be sure to discuss this at the end -->
<!-- as something for future research. -->

@@

Finally, we examine whether flattering advice indeed leads to more positive evaluations of
advisers, or whether flattering advice is dismissed as insencere.
Following our preregistration, we average the ratings on likeability,
warmth,  friendliness, and good-naturedness to create a scale of likeability
($\alpha$ = XXX);
<!-- David: The draft had trustworthiness as part of this scale. -->
<!-- Can we just double-check against the preregistration? -->
<!-- And was Likeable part of that? -->
<!-- We might also want a name for the scale that isn't the same as one -->
<!-- of the items, though I'm not sure what. -->
<!-- Same for trustworthiness -- though that could simply be "Trust" -->
and we create a scale of trustworthiness by averaging the ratings of
trustworthiness and sincerity ($\alpha$ = XXX).

<!-- The Figure is not super clear. How about we try a series of box plots, -->
<!-- with the x-axis as the advised rank? Standard errors will be huge at the -->
<!-- extremes. -->

<!-- Also, we need to do something relative to the advisees' expectations. -->
<!-- I wonder what would be interesting here? Maybe averages for three groups: -->
<!-- more attractive than the participants' own guess, equally attractive, and less -->
<!-- attractive? -->

As seen in Figure \@ref(fig:study2rating), we
found that advisees who suggested that the advisee was more attractive
(lower rank) were indeed rated as more likeable and warm (b = -0.109, p
\< .05; b = -0.133, p \< .001, respectively; See Columns 1 and 2 of Table \@ref(tab:study2regs)).
Interestingly, these benefits are not at the cost of being hypocritical;
advisers who recommend a more favorable rank are viewed as no less trustworthy
(Column 3 of Table \@ref(tab:study2regs)).

We were not powered to do a comparison across the two 
experimental groups and did not preregister such a difference.
<!-- David: But we still need to report warmth & trust across treatments. -->

## Discussion

# General Discussion

Advice substantially shapes people's career and personal outcomes.
Therefore, understanding advisers' motivations has far-reaching
implications. We propose that advisers' focus on advisees' expectations
and the desire to avoid disappointing them can induce gender differences
in the advice given and ultimately in the outcomes achieved. Gender
differences in career outcomes contribute significantly to the gender
wage gap, with women often holding less senior positions and
experiencing worse career outcomes, even with equal performance.
Therefore, identifying a factor such as expectations that contributes to
this gap has important implications for organizational practice.

@@

The findings of this study suggest that organizations can promote gender
equity by calibrating employees' expectations or shifting advisers'
attention away from them. While these measures may not entirely solve
the gender gap problem, they can make a meaningful difference when
unconscious bias interventions have failed to show any effect
[@Chang2019; @Paluck2009]. We believe our findings may be particularly
relevant to societies that value signaling in interpersonal
relationships. Future work
could also examine whether people in different cultures place more
weight on expectations.
